---
title: "The Evolution of Linkin Park: Evaluating Loudness and Instrumentalness Across Time"
author: 
  - Cher Ning-Li
  - Kevin Roe
thanks: "Code and data are available at: https://github.com/cher-ning/spotify-data.git"
date: today
date-format: long
abstract: "This paper evaluates the evolution of Linkin Park's instrumentalness and loudness over their multiple albums. Our results suggest that Linkin Park's instrumentalness increased over time but their loundess remained the same, suggesting that they stayed true to their heavy metal roots. Future analysis should include a qualitative study of Linkin Park to validate our results."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(spotifyr)
library(knitr)

data <- read_rds(here::here("data/02-analysis_data/analysis_data.rds"))

```


# Introduction {#sec-introduction}

Linkin Park is one of the perennial rock banks of our time. Notably, they have gone through multiple changes in their band members and have gone through multiple changes in their sound. In this paper, we will analyze how Linkin Park's instrumentalness and loudness evolved over time, wondering if their sound over time. The remainder of the paper is structured as follows @sec-data-overview discusses the data and the relevant measurement methodologies; @sec-data-results presents the data's results; @sec-discussion discusses our findings. 

# Data Overview {#sec-data-overview}

This dataset on Linkin Park on Spotify was scraped using the Spotify API [@citeSpotify]. We obtained our dataset on Spotify using Spotify Developer Tools [@citespotifyapi]. The variables of interest in the paper are Album Release Date, which shows when the album was released. We're also interested in instrumentalness, which predicts if a track has no vocals, and loudness, which shows how loud a song is. Instrumentalness is measured as a number between 0 and 1, with values closer to 1.0 representing tracks more likely to be purely instrumental with no vocal content. Loudness is measured in decibels averaged across the entire track, with typical ranges between -60 and 0 dB. We believe these two variables characterize key elements of rock music: instrumental solos and loud vocals.

There were 12 rows that only showed the year with no month or date data, so we removed them in our cleaned dataset. 

The paper uses the R programming language [@citeR] to analyze the dataset. The tidyverse [@citetidyverse] and janitor [@citejanitor] packages were used to download and clean the dataset. Finally, ggplot2 [@citeggplot2], [@citetidyverse], and knitr [@citeknitr] packages were used to display the results. 

# Results {#sec-data-results}

```{r}
#| label: fig-instrumentalness-vs-time
#| fig-cap: Linkin Park Album Instrumentalness Across Time
#| echo: false
#| warning: false

data |>
  mutate(album_release_date = ymd(album_release_date)) |>
  ggplot(aes(
    x = album_release_date,
    y = instrumentalness,
  )) +
  geom_jitter(alpha = 0.5, width = 0.3, height = 0) + geom_smooth() +
  theme_minimal() + 
  labs(
    x = "Album release date",
    y = "Instrumentalness of song"
  )
```

@fig-instrumentalness-vs-time shows that instrumentalness gets higher over time. This shows that as Linkin Park's sound evolved over time, the amount of instrumentalness increased. 

\newpage 
```{r}
#| label: fig-loudness-vs-time
#| fig-cap: Linkin Park Album Loudness Across Time
#| echo: false
#| warning: false

data |>
  mutate(album_release_date = ymd(album_release_date)) |>
  ggplot(aes(
    x = album_release_date,
    y = loudness
  )) +
  geom_jitter(alpha = 0.5, width = 0.3, height = 0) + geom_smooth() +
  theme_minimal() + 
  labs(
    x = "Album release date",
    y = "Loudness of song"
  )
```

@fig-loudness-vs-time shows that Linkin Park's loudness in albums did not change as much over time. 

# Discussion {#sec-discussion}

@fig-instrumentalness-vs-time shows that Linkin Park made changes to the extent their songs feature instrumentals than vocals. As time passed, the probability for their songs to contain more instrumetnal features than vocals increased, signalling a change in their music. 

@fig-loudness-vs-time does not suggest that there were a lot of changes in the group's noise. This signals that Linkin Park's sound did not change in terms of stereotypical heavy metal characteristics. Overall, Linkin Park's music changed a lot but certain elements relating to heavy metal rock stayed the same to appeal to their core audience. 

Some limitations of the paper is that it is difficult to quantify how much their music has changed with our variables of interest. A qualitative discussion on Linkin Park's changes is something that needs to be further studied. 

\newpage

\appendix

# Appendix {-}


# Additional data details

## Data Sample {#sec-appendix-sample}

```{r}
#| label: tbl-head
#| tbl-cap: First 5 Rows of Cleaned Data 
#| echo: false
#| warning: false
#| message: false

#### Dataset Head ####
data |>
  slice(1:5) |>
  kable(
    digits = 1,
    booktabs = TRUE,
    linesep = ""
  )

```
\newpage


# References


